On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2018-12-24 manuscripts/IEEE/final_submission.tex
last changes

\begin{abstract}
Learning optimal dictionaries for sparse coding has exposed characteristic sparse features of many natural signals. However, universal conditions guaranteeing the consistency of these learned features in the presence of measurement noise have yet to be stated. Here, we prove very generally that optimal dictionaries and sparse codes are uniquely and stably determined up to measurement error for all diverse enough datasets generated in accordance with the constraints assumed by sparse coding. Applications are given to data analysis, engineering, and neuroscience. 
\end{abstract}

@@

% \cite{sato1975method}
\IEEEPARstart{A}{}common assumption underlying contemporary applications of signal processing and pattern analysis is that each of $N$ observed $n$-dimensional signal samples is a (noisy) linear combination of at most $k$ elementary waveforms drawn from some unknown \emph{dictionary} of size $m$, typically with $m \ll N$ (see \cite{Zhang15} for a comprehensive review of this and related models). Attempts to approximate solutions to this sparsity-constrained inverse problem have provided insight into the structure of many signal classes lacking idiosyncratic formal models (e.g., in vision \cite{wang2015sparse}). In fact, a seminal work in the field demonstrated that dictionaries optimized to sparsely represent small patches of natural images share many qualitative similarities to filters fit to the response properties of simple-cell neurons in mammalian visual cortex \cite{Olshausen96, hurri1996image, bell1997independent, van1998independent}, a major result in theoretical neuroscience. Moreover, these latent waveforms (e.g., ``Gabor'' wavelets) estimated from data appear to be canonical \cite{donoho2001can}; i.e. they emerge in dictionaries learned by a variety of algorithms trained with respect to different natural image datasets.

Motivated by these discoveries and earlier work proposing a related theory of communication between sparsely active neural populations \cite{Isely10}, we address when dictionaries and the sparse representations they induce are uniquely determined by data. Answers to this question have real-world implications. For example, it has been demonstrated that a sparse coding analysis of painting style can detect forgeries \cite{hughes2010}; but this implies all dictionaries consistent with training data generalize in their ability to sparsely encode new samples. Fortunately, several algorithms have recently been proposed which provably recover latent dictionaries under idiosyncratic conditions (see \cite[Sec.~I-E]{Sun16} for a summary of the state-of-the-art). Few theorems, however, can be cited to explain the consistency of inference with respect to this model more broadly; in particular, a universal guarantee of uniqueness and stability in the context of noise has yet to emerge.

Here, we prove very generally that uniqueness and stability of solutions is an expected property of both the sparse coding problem described above as well as a related (and perhaps more well-known) optimization formulation, which instead asks for a dictionary and sparse codes that minimize the mean sparsity over samples.

@@

Our claim is that there exist very general conditions under which any matrix $\mathbf{B}$ that solves this problem is always ``close" to $\mathbf{A}$, and likewise for the sparse codes $\mathbf{\overline x}_i$ and $\mathbf{x}_i$. In fact, provided $n \geq \min(2k,m)$, in almost all cases \ref{InverseProblem} is well-posed (as per Hadamard \cite{Hadamard1902}) given enough data (Cor.~\ref{ProbabilisticCor}). In particular, these explicit, universal guarantees hold without imposing \emph{any} constraints on the recovered matrix $\mathbf{B}$ (e.g. that it, too, satisfy any injectivity criteria), even the number $m$ of dictionary elements it should contain.

Note that any particular solution to Problem \ref{InverseProblem} represents an infinity of equivalent solutions $\mathbf{BPD}$ and $\mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\overline x}_1, \ldots, \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\overline x}_N$ parametrized by a choice of permutation matrix $\mathbf{P}$ and invertible diagonal matrix $\mathbf{D}$ capturing the sparsity-preserving labeling and scaling ambiguities inherent to the problem statement. Previous theoretical work addressing the noiseless case $\eta =0$ (e.g., \cite{li2004analysis, Georgiev05, Aharon06, Hillar15}) with fixed $\overline m = m$ has shown that the solution to Prob.~\ref{InverseProblem}, when it exists, is unique up to these  ambiguities provided the $\mathbf{x}_i$ are sufficiently diverse and the matrix $\mathbf{A}$ satisfies:
\begin{align}\label{SparkCondition}
\mathbf{A}\mathbf{x}_1 = \mathbf{A}\mathbf{x}_2 \implies \mathbf{x}_1 = \mathbf{x}_2, \ \ \ \text{for all $k$-sparse } \mathbf{x}_1, \mathbf{x}_2,
\end{align}
%
also called the \textit{spark condition}, which in any case is necessary to guarantee the uniqueness of arbitrary sequences of $k$-sparse $\mathbf{x}_i$. We generalize these results to the practical setting of additive noise by incorporating the notion of stability with respect to noise.

@@

In the next section, we give precise statements of our main results, which include an explicit form for $\varepsilon(\delta_1, \delta_2)$. We then prove our main theorem (Thm.~\ref{DeterministicUniquenessTheorem}) in Sec.~\ref{DUT} after stating some additional definitions and lemmas required for the proof, including a useful result in combinatorial matrix analysis (Lem.~\ref{MainLemma}, proven in Appendix). We also provide a short argument extending our guarantees to the following more common formulation of the dictionary learning problem as an optimization over model parameters minimizing the total (or equivalently, the mean) number of non-zero entries in the sparse codes (Thm.~\ref{SLCopt}):

\begin{problem}\label{OptimizationProblem}
Find $\mathbf{B}$ and \mbox{$\mathbf{\overline x}_1, \ldots, \mathbf{\overline x}_N \in \mathbb{R}^{\overline m}$} that solve:
\begin{align}\label{minsum}
\min \sum_{i = 1}^N \|\mathbf{\overline x}_i\|_0 \ \ 
\text{subject to} \ \ \|\mathbf{z}_i - \mathbf{B}\mathbf{\overline x}_i\|_2 \leq \eta, \ \text{for all $i$}.
\end{align}
\end{problem}
We then sketch proofs of probabilistic extensions of Thm.~\ref{DeterministicUniquenessTheorem} to  data and dictionaries drawn from probability distributions (Thm.~\ref{robustPolythm} and Cor.~\ref{ProbabilisticCor}). Finally, in Sec.~\ref{Discussion}, we discuss both theoretical and practical applications of our main mathematical findings.

@@

Next, we describe a quantitative spark condition. The \emph{lower bound} of a matrix $\mathbf{M} \in \mathbb R^{n \times m}$ is the largest $\alpha$ with \mbox{$\|\mathbf{M}\mathbf{x}\|_2 \geq \alpha\|\mathbf{x}\|_2$} for all $\mathbf{x} \in \mathbb{R}^m$ \cite{Grcar10}. By compactness of the unit sphere, injective linear maps have nonzero lower bound; hence, if $\mathbf{M}$ satisfies \eqref{SparkCondition}, then each submatrix formed from $2k$ of its columns or less has a strictly positive lower bound.

We generalize the  lower bound to a domain-restricted \emph{union of subspaces} model \cite{vidal2005generalized} derived from a hypergraph $\mathcal{H} \subseteq {[m] \choose k}$. Let $\mathbf{M}_S$ denote the submatrix formed by the columns of $\mathbf{M}$ indexed by $S \subseteq [m]$, with $\mathbf{M}_\emptyset := \mathbf{0}$. (In the sections that follow, we write $\bm{\mathcal{M}}_S$ to denote the column-span of a submatrix $\mathbf{M}_S$, and $\bm{\mathcal{M}}_\mathcal{G}$ to denote $\{\bm{\mathcal{M}}_S\}_{S \in \mathcal{G}}$.) Define: 
%\begin{align*} 
%L_\mathcal{H}(\mathbf{M}) := \min \left\{ \frac{ \|\mathbf{M}(\mathbf{x}_1-\mathbf{x}_2)\|_2 }{ \sqrt{2k} \|\mathbf{x}_1-\mathbf{x}_2\|_2} : \mathbf{x}_1, \mathbf{x}_2 \in \cup_{S \in \mathcal{H}} \bm{\mathcal{M}}_S \right\},
%\end{align*} 
%
%where we write $L_{2k}$ in place of $L_\mathcal{H}$ when $\mathcal{H} = {[m] \choose k}$. Note that if $\mathcal{H}$ covers $[m]$, then $L_2 > L_\mathcal{H}$.\footnote{The reader should beware that $L_2 = L_{[m]}$, whereas $L = L_{\{[m]\}}$. For even $k$, the quantity $1 - \sqrt{k} L_k(\mathbf{M})$ is also known in the compressed sensing literature as the (asymmetric) lower restricted isometry constant \cite{Blanchard2011}.} Clearly, for any $\mathbf{M}$ satisfying \eqref{SparkCondition}, we have $L_{k'}(\mathbf{M}) > 0$ for  $k' \leq 2k$.
%\begin{align*}
%L_\mathcal{H}(\mathbf{M}) := \min \left\{ \frac{ \|\mathbf{M}_S\mathbf{x}\|_2 }{ \sqrt{k} \|\mathbf{x}\|_2} : S \in \mathcal{H} \right\},
%\end{align*} 
%\begin{align*}
%L_\mathcal{H}(\mathbf{M}) := \max \left\{ \alpha: \|\mathbf{M}_S\mathbf{x}\|_2 \geq \alpha \sqrt{k} \|\mathbf{x}\|_2 : S \in \mathcal{H}, \ \ \mathbf{x} \in \mathbb{R}^m \right\},
%\end{align*} 
\begin{align}\label{Ldef}
L_\mathcal{H}(\mathbf{M}) := \frac{1}{\sqrt{k}} \min \left\{ \frac{\|\mathbf{M}_S\mathbf{x}\|_2}{ \|\mathbf{x}\|_2} : S \in \mathcal{H}, \ \ \mathbf{x} \in \mathbb{R}^{|S|} \right\},
\end{align} 
%
where we write $L_{k}$ when $\mathcal{H} = {[m] \choose k}$.\footnote{We note that $1 - \sqrt{k} L_k(\mathbf{M})$ is known as the asymmetric lower restricted isometry constant for matrices $\mathbf{M}$ with unit $\ell_2$-norm columns \cite{Blanchard2011}.} Clearly, $L_{2k}(\mathbf{M}) > 0$ for $\mathbf{M}$ satisfying \eqref{SparkCondition} and $L_{k'}(\mathbf{M}) \geq L_k(\mathbf{M})$ whenever $k' \leq k$. Note also that $L_2 \geq L_{2\mathcal{H}} \geq L_{2k}$ if $\mathcal{H} \subseteq {[m] \choose k}$ has $\cup \mathcal{H} = [m]$.

A vector $\mathbf{x}$ is said to be \emph{supported} in $S \subseteq [m]$ when $\mathbf{x} \in \text{\rm span}\{\mathbf{e}_j: j\in S\}$, where $\mathbf{e}_j$ form the standard column basis in $\mathbb R^m$. 
%Denote by $\mathbf{x}^J$ the subvector formed from the entries of $\mathbf{x}$ indexed by $J$. 
A set of $k$-sparse vectors is said to be in \emph{general linear position} when any $k$ of them are linearly independent. The following is a precise statement of our main result, though we will leave the quantity $C_1 = C_1(\mathbf{A}, \{\mathbf{x}_i\}_{i=1}^N, \mathcal{H})$ undefined until Eq.~\eqref{Cdef1}. All of our theorems assume matrices consist of real numbers.

@@

%We delay defining the explicit constant $C_1$ until Section \ref{DUT} (\eqref{Cdef1}).
%To be clear, the implication of Thm.~\ref{DeterministicUniquenessTheorem} is that $Y = \{\mathbf{Ax}_1, \ldots, \mathbf{Ax}_N\}$ has a stable $k$-sparse representation in $\mathbb{R}^m$, with \eqref{def1} guaranteed provided $\varepsilon$ in Def.~\ref{maindef} does not exceed: 
In words, Thm.~\ref{DeterministicUniquenessTheorem} says that the smaller the difference $\overline m - m$ between assumed and actual latent dictionary sizes, the more columns and coefficients of the original dictionary $\mathbf{A}$ and codes $\mathbf{x}_i$ are contained (up to noise) in the appropriately scaled dictionary $\mathbf{B}$ and codes $\mathbf{\overline x}_i$.  In the particular case when $\overline m = m$, the theorem directly implies that  $Y = \{\mathbf{Ax}_1, \ldots, \mathbf{Ax}_N\}$ has a stable $k$-sparse representation in $\mathbb{R}^m$, with inequalities \eqref{def1} guaranteed for $\varepsilon$ in Def.~\ref{maindef} given by: 
\begin{align}\label{epsdel}
\varepsilon(\delta_1, \delta_2) := \min \left\{ \frac{\delta_1}{ C_1 }, \frac{ \delta_2 L_{2k}(\mathbf{A})}{ 1 + C_1 \left( \delta_2 + \max_{i \in [N]} \|\mathbf{x}_i\|_1  \right) } \right\}.
\end{align}

Note that sparse codes $\mathbf{x}_i$ with a shared support that are in general linear position are straightforward to produce using a ``Vandermonde'' matrix construction (i.e., use the columns of the matrix $[\gamma_{i}^j]_{i,j=1}^{k,N}$, for distinct nonzero $\gamma_i$).  Thus, the assumptions of Thm.~\ref{DeterministicUniquenessTheorem} are easily met, leading to the following direct application to uniqueness in sparse coding. % (Prob.~\ref{InverseProblem}).
%=======
%An immediate practical implication of this result is that there exists a practical procedure to affirm if one's proposed solution $(\mathbf{B}, \mathbf{\overline x}_1, \ldots, \mathbf{x}_N)$ to Prob.~\ref{InverseProblem} is indeed unique (up to noise and inherent ambiguities): simply check that $\mathbf{B}$ and the $\mathbf{\overline x}_i$ satisfy the assumptions on $\mathbf{A}$ and the $\mathbf{x}_i$ in Thm.~\ref{DeterministicUniquenessTheorem}.
%
%%In fact, a more general result (stated clearly in the next section) can be gleaned from our method of proving Thm.~\ref{DeterministicUniquenessTheorem}. Briefly, in cases where $\mathbf{B}$ has $\overline m \neq m$ columns, or $\mathcal{H}$ is not regular or only partially satisfying the SIP, a relation between $\overline m$ and the degree sequence of nodes in $\mathcal{H}$ gives indices $J \subseteq [m]$ defining a submatrix $\mathbf{A}_J$ and subvectors $\mathbf{x}_i^J$ that are recoverable in the sense of \eqref{Cstable} and \eqref{b-PDa}. For example, if $\mathcal{H}$ is $\ell$-regular with the SIP but $m \leq \overline m < m\ell/(\ell - 1)$ then we have nonzero $|J| = \overline m - \ell(\overline m - m)$. The implication here is that the smaller the difference $\overline m - m$, the more columns and code entries of the original $n \times m$ dictionary $\mathbf{A}$ and codes $\mathbf{x}_i$ contained (up to noise) in the appropriately scaled $n \times \overline m$ dictionary $\mathbf{B}$ and codes $\mathbf{\overline x}_i$. When $\overline m = m$, we recover Thm.~\ref{DeterministicUniquenessTheorem}.
%
%%In fact, even if $\mathcal{H}$ is not regular or only partially satisfies the SIP, a relation between $\overline m$ and the degree sequence of nodes in $\mathcal{H}$ may give the indices $J \subseteq [m]$. For sake of brevity, we delay to the next section a clear statement of this more general result.
%
%Regarding the assumptions of Thm.~\ref{DeterministicUniquenessTheorem}, it so happens that sparse codes $\mathbf{x}_i$ in general linear position are straightforward to produce with a ``Vandermonde'' matrix construction \cite{Hillar15}, leading to the following.
%>>>>>>> d1407a275dc391dabad7bb62d3e659f9a4ac4624

@@

Another practical implication of Thm.~\ref{DeterministicUniquenessTheorem} is the following: there is an effective procedure sufficient to affirm if a proposed solution to Prob.~\ref{InverseProblem} is indeed unique (up to noise and inherent ambiguities). Simply check that the matrix and codes satisfy the (computable) assumptions of Thm.~\ref{DeterministicUniquenessTheorem} on $\mathbf{A}$ and the $\mathbf{x}_i$.  
%Another practical implication of Thm.\ref{DeterministicUniquenessTheorem} is the following: there is an effective procedure sufficient to affirm if a proposed solution $(\mathbf{B}, \mathbf{\overline x}_1, \ldots, \mathbf{x}_N)$ to Prob.~\ref{InverseProblem} is indeed unique (up to noise and inherent ambiguities). One simply checks that $\mathbf{B}$ and the $\mathbf{\overline x}_i$ satisfy the respective assumptions of Thm.~\ref{DeterministicUniquenessTheorem}.  
%. on $\mathbf{A}$ and the $\mathbf{x}_i$, respectively.

@@

There are other less direct consequences of Thm.~\ref{DeterministicUniquenessTheorem}.  For instance, we apply it to prove uniqueness and stability of solutions to Prob.~\ref{SLCopt}, the usual optimization problem of interest for those applying dictionary learning to their data.

@@

Thus, choosing the dictionary and codes ``randomly'' almost certainly generates data with a stable sparse representation.

@@

A goal of this work was to explain the emergence of characteristic representations from sparse coding models fit to natural data, despite the varied assumptions underlying the many algorithms in current use. To this end, we have taken an important step toward unifying the great number of publications on the topic by demonstrating very general, deterministic conditions under which identification of parameters in this model is not only possible but also robust to the inevitable uncertainty permeating measurement and model choice.

@@

Beyond an original extension of existing noiseless guarantees \cite{Hillar15} to the noisy regime and their application to Prob.~\ref{OptimizationProblem}, a major innovation in our work is a theory of combinatorial designs for support sets key to the identification of the dictionary. We incorporate this idea into a new fundamental lemma in matrix theory (Lem.~\ref{MainLemma}) that draws upon the definition of a new matrix lower bound induced by a hypergraph. Insights enabled by our combinatorial approach include: 1) a subset of dictionary elements is recoverable even if dictionary size is overestimated, 2) data require only a polynomial number of distinct sparse supports, and 3) the spark condition is not a necessary property of recoverable dictionaries.

>>>>

\begin{abstract}
Learning optimal dictionaries for sparse coding has exposed characteristic sparse features of many natural signals. However, universal conditions guaranteeing the consistency of these learned features in the presence of measurement noise have yet to be stated. Here, we prove very generally that optimal dictionaries and sparse codes are uniquely and stably determined up to measurement error for all diverse enough datasets generated by the sparse coding model. Applications are given to data analysis, engineering, and neuroscience. 
\end{abstract}

@@

% \cite{sato1975method}
\IEEEPARstart{A}{}common assumption underlying contemporary solutions to problems in signal processing and pattern analysis is that each of $N$ observed $n$-dimensional signal samples is a (noisy) linear combination of at most $k$ elementary waveforms drawn from some unknown ``dictionary" of size $m$, typically with $m \ll N$ (see \cite{Zhang15} for a comprehensive review of this and related models). Approximate solutions to this sparsity-constrained inverse problem and its relatives have revealed seemingly characteristic structure in signal classes lacking idiosyncratic formal models (e.g., in vision \cite{wang2015sparse}). Interestingly, a seminal work in the field demonstrated that dictionaries optimized with respect to small patches of natural images share qualitative similarities with filters fit to the response properties of simple-cell neurons in mammalian visual cortex \cite{Olshausen96, hurri1996image, bell1997independent, van1998independent}, a major result in theoretical neuroscience. Moreover, these latent waveforms (e.g., ``Gabor'' wavelets) estimated from data appear to be canonical \cite{donoho2001can}; i.e. they appear in dictionaries learned by a variety of algorithms trained with respect to different natural image datasets.

Motivated by these discoveries and earlier work which proposed a theory of communication between sparsely active neural populations contingent on uniqueness \cite{Isely10}, we address when dictionaries and the sparse representations they induce are uniquely determined by data. Answers to this question have other practical implications. For example, it has been demonstrated that a sparse coding analysis of painting style can detect forgeries \cite{hughes2010}; but this implies all dictionaries consistent with training data generalize in their ability to sparsely encode new samples. Fortunately, several algorithms have recently been proposed which provably recover latent dictionaries under idiosyncratic conditions (see \cite[Sec.~I-E]{Sun16} for a summary of the state-of-the-art). Few theorems, however, can be cited to explain the consistency of inference under this model of data more broadly; in particular, a universal guarantee of uniqueness and stability in the context of noise has yet to emerge.

Here, we prove very generally that uniqueness and stability of solutions is an expected property of the sparse coding problem described above as well as a related (and perhaps more well-known) optimization problem which asks for a dictionary and set of sparse codes that minimize the mean sparsity of the codes over all data samples.

@@

Our claim is that there exist very general conditions under which any matrix $\mathbf{B}$ that solves this problem is always ``close" to $\mathbf{A}$, and likewise for the sparse codes $\mathbf{\overline x}_i$ and $\mathbf{x}_i$. In fact, provided $n \geq \min(2k,m)$, in almost all cases (\ref{InverseProblem}) is well-posed (as per Hadamard \cite{Hadamard1902}) given enough data (Cor.~\ref{ProbabilisticCor}). Importantly, these explicit, universal guarantees apply without the imposition of \emph{any} constraints on the recovered matrix $\mathbf{B}$ (e.g. that it, too, satisfy any injectivity criteria), beyond a cap on the number $m$ of dictionary elements it should contain.

We note that any particular solution to Problem \ref{InverseProblem} actually represents an infinity of equivalent solutions $\mathbf{BPD}$ and $\mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\overline x}_1, \ldots, \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\overline x}_N$ parametrized by a choice of permutation matrix $\mathbf{P}$ and invertible diagonal matrix $\mathbf{D}$ capturing the sparsity-preserving labeling and scaling ambiguities inherent to the problem statement. Previous theoretical work addressing the noiseless case $\eta =0$ (e.g., \cite{li2004analysis, Georgiev05, Aharon06, Hillar15}) with fixed $\overline m = m$ has shown that the solution to Prob.~\ref{InverseProblem}, when it exists, is unique up to these  ambiguities provided the $\mathbf{x}_i$ are sufficiently diverse and the matrix $\mathbf{A}$ satisfies:
\begin{align}\label{SparkCondition}
\mathbf{A}\mathbf{x}_1 = \mathbf{A}\mathbf{x}_2 \implies \mathbf{x}_1 = \mathbf{x}_2, \ \ \ \text{for all $k$-sparse } \mathbf{x}_1, \mathbf{x}_2,
\end{align}
%
also called the \textit{spark condition}, which in any case is necessary to guarantee the uniqueness of arbitrary sequences of $k$-sparse $\mathbf{x}_i$. We generalize these results to the practical setting of additive noise by incorporating a notion of stability with respect to measurement error.

@@

In the next section, we give precise statements of our main results, which include an explicit form for $\varepsilon(\delta_1, \delta_2)$. We then prove our main theorem (Thm.~\ref{DeterministicUniquenessTheorem}) in Sec.~\ref{DUT} after stating some additional definitions and lemmas required for the proof, including a useful result in combinatorial matrix analysis (Lem.~\ref{MainLemma}, proven in Appendix). We also provide an argument extending our guarantees to the following more common formulation of the dictionary learning problem as an optimization over model parameters minimizing the total (or equivalently, the mean) number of non-zero entries in the sparse codes (Thm.~\ref{SLCopt}):

\begin{problem}\label{OptimizationProblem}
Find $\mathbf{B}$ and \mbox{$\mathbf{\overline x}_1, \ldots, \mathbf{\overline x}_N \in \mathbb{R}^{\overline m}$} that solve:
\begin{align}\label{minsum}
\min \sum_{i = 1}^N \|\mathbf{\overline x}_i\|_0 \ \ 
\text{subject to} \ \ \|\mathbf{z}_i - \mathbf{B}\mathbf{\overline x}_i\|_2 \leq \eta, \ \text{for all $i$}.
\end{align}
\end{problem}
We then sketch proofs of extensions of Thm.~\ref{DeterministicUniquenessTheorem} to data and dictionaries drawn from a probability distribution (Thm.~\ref{robustPolythm} and Cor.~\ref{ProbabilisticCor}). Finally, in Sec.~\ref{Discussion}, we discuss both theoretical and practical applications of our main mathematical findings.

@@

Next, we describe a quantitative generalization of the spark condition. The \emph{lower bound} of a matrix $\mathbf{M} \in \mathbb R^{n \times m}$ is the largest $\alpha$ with \mbox{$\|\mathbf{M}\mathbf{x}\|_2 \geq \alpha\|\mathbf{x}\|_2$} for all $\mathbf{x} \in \mathbb{R}^m$ \cite{Grcar10}. By compactness of the unit sphere, injective linear maps have nonzero lower bound; hence, if $\mathbf{M}$ satisfies \eqref{SparkCondition}, then each submatrix formed from $2k$ of its columns or less has a strictly positive lower bound.

We generalize this lower bound to a domain-restricted \emph{union of subspaces} model \cite{vidal2005generalized} derived from a hypergraph $\mathcal{H} \subseteq {[m] \choose k}$. Let $\mathbf{M}_S$ denote the submatrix formed by the columns of $\mathbf{M}$ indexed by $S \subseteq [m]$, with $\mathbf{M}_\emptyset := \mathbf{0}$. (In the sections that follow, we write $\bm{\mathcal{M}}_S$ to denote the column-span of a submatrix $\mathbf{M}_S$, and $\bm{\mathcal{M}}_\mathcal{G}$ to denote $\{\bm{\mathcal{M}}_S\}_{S \in \mathcal{G}}$.) Define: 
%\begin{align*} 
%L_\mathcal{H}(\mathbf{M}) := \min \left\{ \frac{ \|\mathbf{M}(\mathbf{x}_1-\mathbf{x}_2)\|_2 }{ \sqrt{2k} \|\mathbf{x}_1-\mathbf{x}_2\|_2} : \mathbf{x}_1, \mathbf{x}_2 \in \cup_{S \in \mathcal{H}} \bm{\mathcal{M}}_S \right\},
%\end{align*} 
%
%where we write $L_{2k}$ in place of $L_\mathcal{H}$ when $\mathcal{H} = {[m] \choose k}$. Note that if $\mathcal{H}$ covers $[m]$, then $L_2 > L_\mathcal{H}$.\footnote{The reader should beware that $L_2 = L_{[m]}$, whereas $L = L_{\{[m]\}}$. For even $k$, the quantity $1 - \sqrt{k} L_k(\mathbf{M})$ is also known in the compressed sensing literature as the (asymmetric) lower restricted isometry constant \cite{Blanchard2011}.} Clearly, for any $\mathbf{M}$ satisfying \eqref{SparkCondition}, we have $L_{k'}(\mathbf{M}) > 0$ for  $k' \leq 2k$.
%\begin{align*}
%L_\mathcal{H}(\mathbf{M}) := \min \left\{ \frac{ \|\mathbf{M}_S\mathbf{x}\|_2 }{ \sqrt{k} \|\mathbf{x}\|_2} : S \in \mathcal{H} \right\},
%\end{align*} 
%\begin{align*}
%L_\mathcal{H}(\mathbf{M}) := \max \left\{ \alpha: \|\mathbf{M}_S\mathbf{x}\|_2 \geq \alpha \sqrt{k} \|\mathbf{x}\|_2 : S \in \mathcal{H}, \ \ \mathbf{x} \in \mathbb{R}^m \right\},
%\end{align*} 
\begin{align}\label{Ldef}
L_\mathcal{H}(\mathbf{M}) := \frac{1}{\sqrt{k}} \min \left\{ \frac{\|\mathbf{M}_S\mathbf{x}\|_2}{ \|\mathbf{x}\|_2} : S \in \mathcal{H}, \ \ \mathbf{x} \in \mathbb{R}^{|S|} \right\},
\end{align} 
%
where we write $L_{k}$ when $\mathcal{H} = {[m] \choose k}$.\footnote{We note that $1 - \sqrt{k} L_k(\mathbf{M})$ is known as the asymmetric lower restricted isometry constant for matrices $\mathbf{M}$ with unit $\ell_2$-norm columns \cite{Blanchard2011}.} Clearly, $L_{2k}(\mathbf{M}) > 0$ for $\mathbf{M}$ satisfying \eqref{SparkCondition} and $L_{k'}(\mathbf{M}) \geq L_k(\mathbf{M})$ whenever $k' \leq k$. Note also that $L_2 \geq L_{2\mathcal{H}} \geq L_{2k}$ if $\mathcal{H} \subseteq {[m] \choose k}$ has $\cup \mathcal{H} = [m]$.

A vector $\mathbf{x}$ is said to be \emph{supported} in $S \subseteq [m]$ when $\mathbf{x} \in \text{\rm span}\{\mathbf{e}_j: j\in S\}$, where $\mathbf{e}_j$ form the standard column basis in $\mathbb R^m$. 
%Denote by $\mathbf{x}^J$ the subvector formed from the entries of $\mathbf{x}$ indexed by $J$. 
A set of $k$-sparse vectors is said to be in \emph{general linear position} when any $k$ of them are linearly independent.

We are now in a position to give a precise statement of our main result, though we will leave the quantity $C_1 = C_1(\mathbf{A}, \{\mathbf{x}_i\}_{i=1}^N, \mathcal{H})$ undefined until Eq.~\eqref{Cdef1}. All of our theorems assume matrices consist of real numbers.

@@

%We delay defining the explicit constant $C_1$ until Section \ref{DUT} (\eqref{Cdef1}).
%To be clear, the implication of Thm.~\ref{DeterministicUniquenessTheorem} is that $Y = \{\mathbf{Ax}_1, \ldots, \mathbf{Ax}_N\}$ has a stable $k$-sparse representation in $\mathbb{R}^m$, with \eqref{def1} guaranteed provided $\varepsilon$ in Def.~\ref{maindef} does not exceed: 
In words, Thm.~\ref{DeterministicUniquenessTheorem} says that the smaller the difference $\overline m - m$ between the assumed and actual number of latent dictionary elements, the more columns and coefficients of the original dictionary $\mathbf{A}$ and codes $\mathbf{x}_i$ are contained (up to noise) in the appropriately scaled recovered dictionary $\mathbf{B}$ and codes $\mathbf{\overline x}_i$, respectively.  In the particular case when $\overline m = m$, the theorem directly implies that  $Y = \{\mathbf{Ax}_1, \ldots, \mathbf{Ax}_N\}$ has a stable $k$-sparse representation in $\mathbb{R}^m$, with inequalities \eqref{def1} guaranteed for $\varepsilon$ in Def.~\ref{maindef} given by: 
\begin{align}\label{epsdel}
\varepsilon(\delta_1, \delta_2) := \min \left\{ \frac{\delta_1}{ C_1 }, \frac{ \delta_2 L_{2k}(\mathbf{A})}{ 1 + C_1 \left( \delta_2 + \max_{i \in [N]} \|\mathbf{x}_i\|_1  \right) } \right\}.
\end{align}

Note that sparse codes $\mathbf{x}_i$ with a shared support that are in general linear position are straightforward to produce using a ``Vandermonde'' matrix construction (i.e., use the columns of the matrix $[\gamma_{i}^j]_{i,j=1}^{k,N}$, for distinct nonzero $\gamma_i$).  Thus, the assumptions of Thm.~\ref{DeterministicUniquenessTheorem} are easily met. % (Prob.~\ref{InverseProblem}).
%=======
%An immediate practical implication of this result is that there exists a practical procedure to affirm if one's proposed solution $(\mathbf{B}, \mathbf{\overline x}_1, \ldots, \mathbf{x}_N)$ to Prob.~\ref{InverseProblem} is indeed unique (up to noise and inherent ambiguities): simply check that $\mathbf{B}$ and the $\mathbf{\overline x}_i$ satisfy the assumptions on $\mathbf{A}$ and the $\mathbf{x}_i$ in Thm.~\ref{DeterministicUniquenessTheorem}.
%
%%In fact, a more general result (stated clearly in the next section) can be gleaned from our method of proving Thm.~\ref{DeterministicUniquenessTheorem}. Briefly, in cases where $\mathbf{B}$ has $\overline m \neq m$ columns, or $\mathcal{H}$ is not regular or only partially satisfying the SIP, a relation between $\overline m$ and the degree sequence of nodes in $\mathcal{H}$ gives indices $J \subseteq [m]$ defining a submatrix $\mathbf{A}_J$ and subvectors $\mathbf{x}_i^J$ that are recoverable in the sense of \eqref{Cstable} and \eqref{b-PDa}. For example, if $\mathcal{H}$ is $\ell$-regular with the SIP but $m \leq \overline m < m\ell/(\ell - 1)$ then we have nonzero $|J| = \overline m - \ell(\overline m - m)$. The implication here is that the smaller the difference $\overline m - m$, the more columns and code entries of the original $n \times m$ dictionary $\mathbf{A}$ and codes $\mathbf{x}_i$ contained (up to noise) in the appropriately scaled $n \times \overline m$ dictionary $\mathbf{B}$ and codes $\mathbf{\overline x}_i$. When $\overline m = m$, we recover Thm.~\ref{DeterministicUniquenessTheorem}.
%
%%In fact, even if $\mathcal{H}$ is not regular or only partially satisfies the SIP, a relation between $\overline m$ and the degree sequence of nodes in $\mathcal{H}$ may give the indices $J \subseteq [m]$. For sake of brevity, we delay to the next section a clear statement of this more general result.
%
%Regarding the assumptions of Thm.~\ref{DeterministicUniquenessTheorem}, it so happens that sparse codes $\mathbf{x}_i$ in general linear position are straightforward to produce with a ``Vandermonde'' matrix construction \cite{Hillar15}, leading to the following.
%>>>>>>> d1407a275dc391dabad7bb62d3e659f9a4ac4624

@@

A practical implication of Thm.~\ref{DeterministicUniquenessTheorem} is the following: there is an effective procedure sufficient to affirm if a proposed solution to Prob.~\ref{InverseProblem} is indeed unique (up to noise and inherent ambiguities). One need simply to check that the matrix and codes satisfy the (computable) assumptions of Thm.~\ref{DeterministicUniquenessTheorem} on $\mathbf{A}$ and the $\mathbf{x}_i$.  
%Another practical implication of Thm.\ref{DeterministicUniquenessTheorem} is the following: there is an effective procedure sufficient to affirm if a proposed solution $(\mathbf{B}, \mathbf{\overline x}_1, \ldots, \mathbf{x}_N)$ to Prob.~\ref{InverseProblem} is indeed unique (up to noise and inherent ambiguities). One simply checks that $\mathbf{B}$ and the $\mathbf{\overline x}_i$ satisfy the respective assumptions of Thm.~\ref{DeterministicUniquenessTheorem}.  
%. on $\mathbf{A}$ and the $\mathbf{x}_i$, respectively.

@@

There are other less direct consequences of Thm.~\ref{DeterministicUniquenessTheorem}.  For instance, we use it to prove uniqueness and stability of solutions to Prob.~\ref{SLCopt}, the usual optimization problem of interest for those applying dictionary learning to their data.

@@

Thus, drawing the dictionary and sparse codes from any continuous probability distribution almost certainly generates data with a stable sparse representation.

@@

A motivation for this work was the emergence of seemingly characteristic representations from sparse coding models fit to natural data, despite the varied assumptions underlying the many algorithms in current use. To this end, we have taken an important step toward unifying the great number of publications on the topic by demonstrating very general, deterministic conditions under which identification of parameters in this model is not only possible but also robust to the inevitable uncertainty permeating measurement and model choice.

@@

Beyond an original extension of existing noiseless guarantees \cite{Hillar15} to the noisy regime and their application to Prob.~\ref{OptimizationProblem}, a major innovation in our work is a theory of combinatorial designs for support sets key to the identification of the dictionary. We incorporate this idea into a fundamental lemma in matrix theory (Lem.~\ref{MainLemma}) that draws upon the definition of a new matrix lower bound induced by a hypergraph. Insights enabled by our combinatorial approach include: 1) a subset of dictionary elements is recoverable even if dictionary size is overestimated, 2) data require only a polynomial number of distinct sparse supports, and 3) the spark condition is not a necessary property of recoverable dictionaries.
